First Order Alternation

### 2018-09-18 quantifiers/draft.tex
ok

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over $\Data$. Terms are either constants of sort
$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first-order
formulae is defined by the syntax below:
\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
$\phi_1 \vee \phi_2$, $\phi_1 \rightarrow \phi_2$ and $\forall x ~.~
\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
$\fv{}{\phi}$ the set of free variables in $\phi$. A \emph{sentence}
is a formula $\phi$ in which each variable occurs under the scope of a
quantifier, i.e. $\fv{}{\phi} = \emptyset$. A formula is
\emph{positive} if each predicate symbol occurs under an even number
of negations. The size $\len{\phi}$ of a formula $\phi$ is the number
of symbols needed to write it down.

@@

An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
obtained by replacing each variable $x$ by the value $\nu(x)$ and
evaluating each function application. For a formula $\phi$, we define
the forcing relation $\I,\nu \models \phi$ recursively on the
structure of $\phi$, as follows:
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
and behaves like $\nu$ elsewhere. For a formula $\phi$ and a valuation
$\nu$, we define $\sem{\phi}_\nu \isdef \set{\I \mid \I,\nu \models
  \phi}$ and drop the $\nu$ subscript for sentences. A sentence $\phi$
is \emph{satisfiable} (\emph{unsatisfiable}) if
$\sem{\phi}\neq\emptyset$ ($\sem{\phi}=\emptyset$). An element of
$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula $\phi$ is
\emph{valid} if $\I,\nu \models \phi$ for every interpretation $\I$
and every valuation $\nu$.

@@

%% \[\accform{\alpha} = \stamp{\iota}{0} \wedge 
%% \bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
%%   \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
%% \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
%% \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
%% ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot\]

@@

\begin{example}\label{ex:quant-elim}
  Consider the automaton $\A = \tuple{\set{a_1,a_2}, \set{x}, \set{q},
    \iota, \set{q}, \Delta}$, where:
  \[\begin{array}{rcl}
  \iota & = & \exists z ~.~ q(z) \wedge z \geq 0 \\
  \Delta & = & \set{q(y) \arrow{a_1(x)}{} \forall z ~.~ z \leq x+y \rightarrow q(y-z+x),~
    q(y) \arrow{a_2(x)}{} \exists z ~.~ z \geq x-y \wedge q(y+z+x)}
  \end{array}\]
  For the input event sequence $\alpha = a_1a_2$, the path and acceptance formulae are:
  \[\begin{array}{rcl}
  \pathform{\alpha} = \accform{\alpha} & = & \exists z ~.~ \stamp{q}{0}(z) \wedge z \geq 0 \\
  && \forall y ~.~ \stamp{q}{0}(y) \rightarrow (\forall z ~.~ z\leq\stamp{x}{1}+y \rightarrow \stamp{q}{1}(y-z+\stamp{x}{1})) \\
  && \forall y ~.~ \stamp{q}{1}(y) \rightarrow \exists z ~.~ z\geq\stamp{x}{2}-y \wedge \stamp{q}{2}(y+z+\stamp{x}{2})
  \end{array}\]
  The result of eliminating the path quantifiers, in prenex normal form, is shown below:
  \[\begin{array}{rcl}
  \quantform{\alpha} & = & \exists z_1\forall z_2\exists z_3 ~.~ \stamp{q}{0}(z_1) \wedge z_1 \geq 0 \\
  && \wedge~ [\stamp{q}{0}(z_1) \rightarrow (z_2 \leq \stamp{x}{1}+z_1 \rightarrow \stamp{q}{1}(z_1-z_2+\stamp{x}{1}))] \\
  && \wedge~ [\stamp{q}{1}(z_1-z_2+\stamp{x}{1}) \rightarrow z_3 \geq \stamp{x}{2} -z_1+z_2-\stamp{x}{1} \wedge 
    \stamp{q}{2}(z_1-z_2+\stamp{x}{1}+z_3+\stamp{x}{2})] \enspace\enspace\hfill\blacksquare
  \end{array}\]
\end{example}

@@

\begin{example}[Contd. from Example \ref{ex:quant-elim}]\label{ex:pred-elim}
  The result of the elimination of predicate atoms from the acceptance
  formula in Example \ref{ex:quant-elim} is shown below:
  \[\exists z_1 \forall z_2 \exists z_3 ~.~ (z_2 \leq \stamp{x}{1}+z_1 
  \rightarrow z_3 \geq \stamp{x}{2} -z_1+z_2-\stamp{x}{1}) \wedge z_1
  \geq 0\] Since this formula is satisfiable, by Lemma
  \ref{lemma:quant-pred-acceptance} below, there exists a word $w$
  with input event sequence $\event{w} = a_1a_2$ that is accepted by the
  automaton $\A$ from Example
  \ref{ex:quant-elim}. \hfill$\blacksquare$
\end{example}

@@

  \begin{fact}
    Given formulae $\phi$ and $\psi$, such that $\phi$ is positive
    $q(t_1,\ldots,t_{\#(q)})$ is the only one occurrence of the
    predicate symbol $q$ in $\phi$ and no predicate atom with
    predicate symbol $q$ occurs in $\psi(y_1,\ldots,y_{\#(q)})$, for
    each interpretation $\I$ and each valuation $\nu$, we have:
    \[\begin{array}{l} 
    \I,\nu \models \phi \wedge q(t_1,\ldots,t_{\#(q)}) \rightarrow
    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}] \iff \\ \nu \models
    \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]\enspace.
    \end{array}\]
  \end{fact}
  \proof{ We assume w.l.o.g. that $\phi$ is quantifier-free. The proof
    can be easily generalized to the case $\phi$ has quantifiers.

@@

The problem of checking emptiness of a given automata is undecidable,
even for automata with predicates of arity two, whose transition rules
use only equalities and disequalities, with no transition quantifiers
\cite{Farzan15}. Since even such simple classes of alternating
automata have no general decision procedure, we use an
abstraction-refinement semi-algorithm based on \emph{lazy annotation}
\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
procedure systematically explores the set of execution traces of a
program in search for a counterexample. When a spurious counterexample
is found, the search backtracks and the program path is annotated with
a set of learned facts, that excludes the path from future
searches. These facts are interpolants extracted from the proof of
spuriousness of the given (finite) path.

Typically, when checking the unreachability of a set of program
configurations (or, more generally, the infeasability of a set of Horn
clauses \cite{BjornerGurfinkelMcMillanRybalchenko15}), the
interpolants used to annotate the unfolded control structure are
assertions about the values of the program variables in a given
control state $q$, at a certain step $i \geq 0$ of a given
execution. As we discuss next, it is useful to distinguish between
\begin{inparaenum}[(i)]
\item locality of interpolants w.r.t. a given control state and
%
\item locality w.r.t. a given time stamp.
\end{inparaenum} 
In logical terms, \emph{control-local} interpolants are defined by
formulae involving a single predicate symbol, whereas
\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
and variables $\stamp{x}{i}$, for a single $i \geq 0$.

\paragraph{Remark}
When considering an alternating model of computation, control-local
interpolants are not enough to prove emptiness, because of the
synchronization of several branches of the computation on the same
sequence of input values. Consider, for instance, an automaton with
the following transition rules and final state $q_f$:
\[\begin{array}{rclcrcl}
q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
q_1(y) \arrow{a(x)}{} y+x > 0 \wedge q_f \\
q_1(y) & \arrow{a(x)}{} & q_1(y+x) && q_2(y) \arrow{a(x)}{} y-x > 0 \wedge q_f \\
q_2(y) & \arrow{a(x)}{} & q_2(y-x)
\end{array}\]
Started in an initial configuration $q_0(0)$ with an input word
$(a,\nu_1) \ldots (a,\nu_{n-1}) (b,\nu_n)$, such that $\nu_i(x)=k_i$,
the automaton executes as follows:
\[\begin{array}{l}
q_0(0) \arrow{(a,\nu_1)}{} q_1(k_1)
\wedge q_2(-k_1) \ldots \arrow{(a,\nu_{n-1})}{} 
q_1(\sum_{i=1}^{n-1} k_i) \wedge q_2(-\sum_{i=1}^{n-1} k_i) \arrow{(a,\nu_n)}{} \\
\sum_{i=1}^{n} k_i > 0 \wedge \sum_{i=1}^{n} k_i < 0 \wedge q_f \equiv \bot
\end{array}\] 
An overapproximation of the set of cubes generated after one or more
steps is defined by the formula: $\exists x_1 \exists x_2 ~.~ q_1(x_1)
\wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a formula using
one occurrence of a predicate would give a too rough overapproximation
of this set, unable to prove the emptiness of the
automaton. \hfill$\blacksquare$

First, let us give the formal definition of the class of interpolants
we shall work with. Given a formula $\phi$, the \emph{vocabulary} of
$\phi$ is the set of uninterpreted function symbols, predicate symbols
$q \in \stamp{Q}{i}$ and variables $x \in \stamp{X}{i}$, occurring in
$\phi$, for some $i\geq0$. Observe that quantified variables and the
interpreted function symbols of the data theory\footnote{E.g.,\ the
  arithmetic operators of addition and multiplication, when $\Data$ is
  the set of integers.} do not belong to the vocabulary of a
formula. By $\pset{+}{\phi}$ [$\pset{-}{\phi}$] we denote the set of
predicate symbols that occur in $\phi$ under an even [odd] number of
negations.

Given formulae $\phi$ and $\psi$ such that $\phi \wedge \psi$ is
unsatisfiable, a \emph{Lyndon interpolant} \cite{Lyndon59} is a
first-order formula $I$ such that: \begin{compactitem}
\item the formula $\phi \rightarrow I$ is valid,
\item the formula $I \wedge \psi$ is unsatisfiable, 
\item $\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, 
\item $\pset{+}{I} \subseteq \pset{+}{\phi} \cap \pset{+}{\psi}$ and
  $\pset{-}{I} \subseteq \pset{-}{\phi} \cap \pset{-}{\psi}$.
\end{compactitem}
The existence of Lyndon interpolants for first-order logic was proved
by a non-constructive model-theoretic argument \cite{Lyndon59}. In the
following, we shall consider only theories of the data domain $\Data$,
for which there exists a decision procedure for satisfiability, able
to produce Lyndon interpolants. For instance, interpolating decision
procedures exist for linear real and integer arithmetic, with
uninterpreted function symbols (UFLRA and UFLIA)
\cite{RybalchenkoSofronieStokkermans}. In the rest of this section, we
fix an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$.

\begin{definition}\label{def:generalized-lyndon-interpolant}
  Given a non-empty sequence of input events $\alpha = a_1 \ldots a_n
  \in \Sigma^*$, a \emph{generalized Lyndon interpolant} is a sequence
  $I_0,I_1,\ldots,I_n$ of formulae such that, for all $k =
  0,\ldots,n-1$: \begin{compactenum}
    \item $\voc{I_k} = \pset{+}{I_k}$ and $\pset{-}{I_k} = \emptyset$, 
      %
    \item $\iota \models I_0$ and 
      \(I_k \wedge \left(\bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi}
      \forall \vec{y} ~.~ q(\vec{y}) \rightarrow
      \nextf{\psi}\right) \models \nextf{I}_{k+1}\), 
      %
    \item the formula $I_n \wedge \bigwedge_{q \in Q \setminus F}
      \forall \vec{y} ~.~ q(\vec{y}) \rightarrow \bot$ is
      unsatisfiable.
  \end{compactenum}
\end{definition}

\comment[ri]{I am here}

The solution we adopt is a generalization of the classical approach of
extracting the interpolants from the proof of unsatisfiability of a
time-stamped path formula. Given a sequence of input events $\alpha$,
we consider the acceptance formula $\accform{\alpha}$. If this formula
is unsatisfiable, we show the existence of a sequence of (possibly
quantified) positive interpolants that overapproximate the set of
cubes generated at each step during an execution over a word $w$,
whose even sequence is $\event{w} = \alpha$. Moreover, these
interpolants are local, in the sense that each formula uses only the
values of the variables and the state predicates stamped with its
index in the sequence.

Unlike with the classical path formulae for integer programs
\cite{McMillan06} and Horn clauses
\cite{BjornerGurfinkelMcMillanRybalchenko15}, that are expressed using
linear or Presburger arithmetic, the existence of local interpolants
for theories using first-order predicates relies on a non-constructive
argument, the Lyndon Interpolation Theorem \cite{Lyndon59}. Since
usual decidable theories, such as Presburger arithmetic, become
undecidable in the presence of monadic predicate symbols
\cite{Halpern91}, we cannot hope to find an automated interpolant
generation method, in general.

But here is where the predicate-free path formulae (Definition
\ref{def:substform}) come into play. For a given event sequence
$\alpha$, the automaton $\A$ accepts a word $w$ such that $\event{w} =
\alpha$ if and only if $\substform{\alpha}$ is satisfiable (Lemma
\ref{lemma:substitution}). Assuming further that the equality atoms in
the transition rules of $\A$ are written in the language of a
decidable first-order theory, such as Presburger arithmetic, Lemma
\ref{lemma:substitution} gives us an effective way of checking
emptiness of $\A$ relative to a given event sequence. However, this
method does not cope well with lazy annotation, because there is no
way to extract, from the unsatisfiability proof of
$\substform{\alpha}$, the interpolants needed to annotate
$\alpha$. This is because \begin{inparaenum}[(i)]
\item the formula $\substform{\alpha}$, obtained by repeated
  substitutions (Definition \ref{def:substform}) loses track of the
  execution semantics of $\A$, and 
%
\item quantifiers may occur nested in $\substform{\alpha}$, making it
  difficult to write $\substform{\alpha}$ as a proper interpolation
  problem.
\end{inparaenum}

The solution we adopt in this paper relies on an additional assumption
related to the decidability of the predicate-free theory in which
$\substform{\alpha}$ formulae are written, namely that the theory has
an effective \emph{witness-providing quantifier elimination}
procedure. More precisely, we assume that, for each formula $\exists x
~.~ \phi(x)$, there exists an effectively computable term $t$, in
which $x$ does not occur, such that $\exists x ~.~ \phi$ and
$\phi[t/x]$ are equi-satisfiable. These terms, called \emph{witnesses}
in the following, are actual interpretations of the Skolem function
symbols from the following folklore theorem:

>>>>

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over $\Data$. Terms are either constants of sort
$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first-order
formulae is defined by the syntax below:
\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
$\phi_1 \vee \phi_2$, $\phi_1 \rightarrow \phi_2$ and $\forall x ~.~
\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
$\fv{}{\phi}$ the set of free variables in $\phi$. The size
$\len{\phi}$ of a formula $\phi$ is the number of symbols needed to
write it down.

A \emph{sentence} is a formula $\phi$ in which each variable occurs
under the scope of a quantifier, i.e. $\fv{}{\phi} = \emptyset$.  A
formula is \emph{positive} if each predicate symbol occurs under an
even number of negations. A formula is in \emph{prenex form} if it is
of the form $\varphi = Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$
has no quantifiers. In this case we call $\phi$ the \emph{matrix} of
$\varphi$. Every first-order formula can be written in prenex form, by
renaming each quantified variable to a unique name and moving the
quantifiers upfront.

@@

An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
obtained by replacing each variable $x$ by the value $\nu(x)$ and
evaluating each function application. For a formula $\phi$, we define
the forcing relation $\I,\nu \models \phi$ recursively on the
structure of $\phi$, as follows:
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
and behaves like $\nu$ elsewhere. For a formula $\phi$ and a valuation
$\nu$, we define $\sem{\phi}_\nu \isdef \set{\I \mid \I,\nu \models
  \phi}$ and drop the $\nu$ subscript for sentences. A sentence $\phi$
is \emph{satisfiable} (\emph{unsatisfiable}) if
$\sem{\phi}\neq\emptyset$ ($\sem{\phi}=\emptyset$). An element of
$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula $\phi$ is
\emph{valid} if $\I,\nu \models \phi$ for every interpretation $\I$
and every valuation $\nu$. For two formulae $\phi$ and $\psi$ we write
$\phi \models \psi$ for $\sem{\phi} \subseteq \sem{\psi}$, in which
case we say that $\phi$ \emph{entails} $\psi$.

@@

@@

\begin{example}\label{ex:quant-elim}
  Consider the automaton $\A = \tuple{\set{a_1,a_2}, \set{x},
    \set{q,q_f}, \iota, \set{q_f}, \Delta}$, where:
  \[\begin{array}{rcl}
  \iota & = & \exists z ~.~ z \geq 0 \wedge q(z) \\
  \Delta & = & \set{
    q(y) \arrow{a_1(x)}{} x\geq 0 \wedge \forall z ~.~ z \leq y \rightarrow q(x+z),~
    q(y) \arrow{a_2(x)}{} y<0 \wedge q_f(x+y)
  }
  \end{array}\]
  For the input event sequence $\alpha = a_1a_2$, the acceptance
  formula is:
  \[\begin{array}{rcl}
  \accform{\alpha} & = & \exists z ~.~ z \geq 0 \wedge \stamp{q}{0}(z) ~\wedge \\
  && \forall y ~.~ \stamp{q}{0}(y) \rightarrow [\stamp{x}{1}\geq0 \wedge \forall z ~.~ z\geq y 
    \rightarrow \stamp{q}{1}(\stamp{x}{1}+z)] ~\wedge \\
  && \forall y ~.~ \stamp{q}{1}(y) \rightarrow [y<0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+y)]
  \end{array}\]
  The result of eliminating the path quantifiers, in prenex normal form, is shown below:
  \[\begin{array}{rcl}
  \quantform{\alpha} & = & \exists z_1\forall z_2 ~.~ z_1 \geq 0 \wedge \stamp{q}{0}(z_1) ~\wedge \\
  && [\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_2))] ~\wedge \\
  && [\stamp{q}{1}(\stamp{x}{1}+z_2) \rightarrow \stamp{x}{1}+z_2 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_2)] 
  \enspace\enspace\hfill\blacksquare
  \end{array}\]
\end{example}

@@

\begin{example}[Contd. from Example \ref{ex:quant-elim}]\label{ex:pred-elim}
  The result of the elimination of predicate atoms from the acceptance
  formula in Example \ref{ex:quant-elim} is shown below:
  \[\substform{\alpha} = 
  \exists z_1 \forall z_2 ~.~ z_1 \geq 0 \wedge 
          [\stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow
            \stamp{x}{1}+z_2 < 0)]\] Since this formula is
          unsatisfiable, by Lemma \ref{lemma:quant-pred-acceptance}
          below, no word $w$ with input event sequence $\event{w} =
          a_1a_2$ is accepted by the automaton $\A$ from Example
          \ref{ex:quant-elim}. \hfill$\blacksquare$
\end{example}

@@

  \begin{fact}\label{fact:subst}
    Given formulae $\phi$ and $\psi$, such that $\phi$ is positive
    $q(t_1,\ldots,t_{\#(q)})$ is the only one occurrence of the
    predicate symbol $q$ in $\phi$ and no predicate atom with
    predicate symbol $q$ occurs in $\psi(y_1,\ldots,y_{\#(q)})$, for
    each interpretation $\I$ and each valuation $\nu$, we have:
    \[\begin{array}{l} 
    \I,\nu \models \phi \wedge q(t_1,\ldots,t_{\#(q)}) \rightarrow
    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}] \iff \\ \nu \models
    \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]\enspace.
    \end{array}\]
  \end{fact}
  \proof{ We assume w.l.o.g. that $\phi$ is quantifier-free. The proof
    can be easily generalized to the case $\phi$ has quantifiers.

@@

The problem of checking emptiness of a given automata is undecidable,
even for automata with predicates of arity two, whose transition rules
use only equalities and disequalities, with no transition quantifiers
\cite{Farzan15}. Since even such simple classes of alternating
automata have no general decision procedure for emptiness, we use an
abstraction-refinement semi-algorithm based on \emph{lazy annotation}
\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
procedure systematically explores the set of execution traces of a
program in search for a counterexample. When a spurious counterexample
is found, the search backtracks and the program path is annotated with
a set of learned facts, that excludes the path from future
searches. These facts are interpolants extracted from the proof of
spuriousness of the given (finite) path. The semi-algorithm uses
moreover a coverage relation to ensure that a spurious counterexample
is never explore twice, thus making progress towards either finding a
real counterexample or a proof of emptiness. In the rest of this
section, we discuss the generation of interpolants
(\S\ref{sec:interpolants}) and the coverage (\S\ref{sec:coverage})
used by our emptiness checking semi-algorithm (\S\ref{sec:impact}).

\subsection{Interpolants}\label{sec:interpolants}

Typically, when checking the unreachability of a set of program
configurations or, more generally, the infeasability of a set of Horn
clauses \cite{BjornerGurfinkelMcMillanRybalchenko15}, the interpolants
used to annotate the unfolded control structure are assertions about
the values of the program variables in a given control state, at a
certain step of a given execution. As we discuss next, it is useful to
distinguish between
\begin{inparaenum}[(i)]
\item locality of interpolants w.r.t. a given control state (control
  locality) and
%
\item locality w.r.t. a given time stamp (time locality).
\end{inparaenum} 
In logical terms, \emph{control-local} interpolants are defined by
formulae involving a single predicate symbol, whereas
\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
and variables $\stamp{x}{i}$, for a single $i \geq 0$.

\paragraph{Remark}
When considering an alternating model of computation, control-local
interpolants are not always enough to prove emptiness, because of the
synchronization of several branches of the computation on the same
sequence of input values. Consider, for instance, an automaton with
the following transition rules and final state $q_f$:
\[\begin{array}{rclcrcl}
q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
q_1(y) \arrow{a(x)}{} y+x > 0 \wedge q_f \\
q_1(y) & \arrow{a(x)}{} & q_1(y+x) && q_2(y) \arrow{a(x)}{} y-x > 0 \wedge q_f \\
q_2(y) & \arrow{a(x)}{} & q_2(y-x)
\end{array}\]
Started in an initial configuration $q_0(0)$ with an input word
$(a,\nu_1) \ldots (a,\nu_{n-1}) (a,\nu_n)$, such that $\nu_i(x)=k_i$,
the automaton executes as follows:
\[\begin{array}{l}
q_0(0) \arrow{(a,\nu_1)}{} q_1(k_1)
\wedge q_2(-k_1) \ldots \arrow{(a,\nu_{n-1})}{} 
q_1(\sum_{i=1}^{n-1} k_i) \wedge q_2(-\sum_{i=1}^{n-1} k_i) \arrow{(a,\nu_n)}{} \\
\sum_{i=1}^{n} k_i > 0 \wedge \sum_{i=1}^{n} k_i < 0 \wedge q_f \equiv \bot
\end{array}\] 
An overapproximation of the set of cubes generated after one or more
steps is defined by the formula: $\exists x_1 \exists x_2 ~.~ q_1(x_1)
\wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a control-local
formula using one occurrence of a predicate would give a too rough
overapproximation of this set, unable to prove the emptiness of the
automaton. \hfill$\blacksquare$

First, let us give the formal definition of the class of interpolants
we shall work with. Given a formula $\phi$, the \emph{vocabulary} of
$\phi$, denoted $\voc{\phi}$ is the set of predicate symbols $q \in
\stamp{Q}{i}$ and variables $x \in \stamp{X}{i}$, occurring in $\phi$,
for some $i\geq0$. For a term $t$, its vocabulary $\voc{t}$ is the set
of variables that occur in $t$. Observe that quantified variables and
the interpreted function symbols of the data
theory\footnote{E.g.,\ the arithmetic operators of addition and
  multiplication, when $\Data$ is the set of integers.}  do not belong
to the vocabulary of a formula. By $\pset{+}{\phi}$ [$\pset{-}{\phi}$]
we denote the set of predicate symbols that occur in $\phi$ under an
even [odd] number of negations.

\begin{definition}[\cite{Lyndon59}]\label{def:lyndon-interpolant}
Given formulae $\phi$ and $\psi$ such that $\phi \wedge \psi$ is
unsatisfiable, a \emph{Lyndon interpolant} is a formula $I$ such that
$\phi \models I$, the formula $I \wedge \psi$ is unsatisfiable,
$\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, $\pset{+}{I} \subseteq
\pset{+}{\phi} \cap \pset{+}{\psi}$ and $\pset{-}{I} \subseteq
\pset{-}{\phi} \cap \pset{-}{\psi}$.
\end{definition}
The existence of Lyndon interpolants for first-order logic, with no
predefined interpretation of function symbols\footnote{A model
  specifies the universe and the interpretation of function symbols.},
was proved by a non-constructive model-theoretic argument
\cite{Lyndon59}. In the following, we shall consider only those
theories of the data domain $\Data$, for which there exists a decision
procedure for satisfiability, able to produce Lyndon interpolants. For
instance, such interpolating decision procedures exist for linear real
and integer arithmetic, with uninterpreted function symbols (UFLRA and
UFLIA) \cite{RybalchenkoSofronieStokkermans}. In the rest of this
section, let us fix an automaton $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$.

\begin{definition}\label{def:generalized-lyndon-interpolant}
  Given a non-empty sequence of input events $\alpha = a_1 \ldots a_n
  \in \Sigma^*$, a \emph{generalized Lyndon interpolant (GLI)} is a
  sequence $(I_0,\ldots,I_n)$ of formulae such that, for all $k \in
  [n-1]$: \begin{compactenum}
    \item $\pset{-}{I_k} = \emptyset$,
      %
    \item $\stamp{\iota}{0} \models I_0$ and \(I_k \wedge
      \Big(\bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta}
      \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
      \rightarrow \stamp{\psi}{k+1}\Big) \models I_{k+1}\),
      %
    \item $I_n \wedge \bigwedge_{q \in Q \setminus F} \forall y_1
      \ldots \forall y_{\#(q)} ~.~ q(\vec{y})$ is unsatisfiable.
  \end{compactenum}
  Moreover, the GLI is \emph{local} iff $\voc{I_k} \cap \stamp{Q}{\leq n}
  \subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap \stamp{X}{\leq n}
  \subseteq \stamp{X}{k}$, for all $k \in [n]$.
\end{definition}
The following proposition states the existence of local GLI for the
theories in which Lyndon's Interpolation Theorem holds.

\begin{proposition}\label{prop:local-gli}
  If there exists a Lyndon interpolant for any two formulae $\phi$ and
  $\psi$, such that $\phi \wedge \psi$ is unsatisfiable, then any
  sequence of input events $\alpha = a_1 \ldots a_n \in \Sigma^*$,
  such that $\accform{\alpha}$ is unsatisfiable, has a local GLI
  $(I_0,\ldots,I_n)$.
\end{proposition}
\proof{ By definition, $\accform{\alpha}$ is the formula:
  \[\stamp{\iota}{0} \wedge 
 \bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
   \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
 \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
 \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
 ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot\]
 We define the formulae: 
 \[\begin{array}{rcl}
 \varphi_i & \isdef & \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
 \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i},~ \text{for all}~ i \in [1,n] \\
 \psi & \isdef & \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
 ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
 \end{array}\]
 Observe that $\voc{\stamp{\iota}{0}} \subseteq \stamp{Q}{0}$,
 $\voc{\varphi_i} \subseteq
 \stamp{Q}{i-1}\cup\stamp{Q}{i}\cup\stamp{X}{i}$, for all $i \in
       [1,n]$, and $\voc{\psi} \subseteq \stamp{Q}{n}$.  We apply
       Lyndon's Interpolation Theorem for the formulae
       $\stamp{\iota}{0}$ and $\bigwedge_{i=1}^n \varphi_i \wedge
       \psi$ and obtain a formula $I_0$, such that $\stamp{\iota}{0}
       \models I_0$, $I_0 \wedge \bigwedge_{i=1}^n \varphi_i \wedge
       \psi$ is unsatisfiable, $\voc{I_0} \subseteq
       \voc{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n \voc{\varphi_i}
       \cup \voc{\psi}) \subseteq \stamp{Q}{0}$ and $\pset{-}{I_0}
       \subseteq \pset{-}{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n
       \pset{-}{\varphi} \cup \pset{-}{\psi}) = \emptyset$. Repeating
       the reasoning for the formulae $I_0 \wedge \varphi_1$ and
       $\bigwedge_{i=2}^n \varphi_i \wedge \psi$, we obtain $I_1$,
       such that $I_0 \wedge \varphi_1 \models I_1$, $I_1 \wedge
       \bigwedge_{i=2}^n \varphi_i \wedge \psi$ is unsatisfiable,
       $\voc{I_1} \subseteq (\voc{I_0} \cup \voc{\varphi_1}) \cap
       (\bigcup_{i=2}^n \voc{\varphi_i} \cup \voc{\psi}) \subseteq
       \stamp{Q}{1}$ and $\pset{-}{I_1} \subseteq (\pset{-}{I_0} \cup
       \pset{-}{\varphi_1}) \cap (\bigcup_{i=2}^n \pset{-}{\varphi_i}
       \cup \pset{-}{\psi}) = \emptyset$. Continuing in this way, we
       obtain formulae $I_0, I_1, \ldots, I_{n}$ as required. \qed}
